---
title: "inst_627_suarez_final"
output: html_document
date: "2023-11-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is the R Markdown file for Pablo Suarez's final project in INST627: DATA ANALYTICS FOR INFORMATION PROFESSIONALS.

In this file, I am analyzing gender pay disparities in the United States. As mentioned in my project proposal, this is an issue that has persisted for decades despite intervention from legislators and recognition among the general public. U.S. Census data suggests that it wasn’t until 1990 that women reached a threshold of earning 70 cents to every dollar earned by men. Almost 30 years later, the gap has closed to roughly 84 cents for every dollar men earn. 

To better understand this issue, I would like to answer the following research questions: 
1. How does the gender pay disparity vary by race, age and education?
2. Which industries have the most and least significant wage discrepancies?
3. At what rate do men and women participate in the workforce?
4. (Potential) Highlight major policies or economic events that have contributed or shaped these trends.


```{r}
# Turn off scientific notation
options(scipen=999)

#load libraries
library(tidyverse)
library(janitor)
library(readr)
library(lubridate)
library(refinr)
library(dplyr)
library(ggthemes)
library(ggplot2)
```






2. Which industries have the most and least significant wage discrepancies?

```{r}

largest_industry_gaps <- read_csv("data/largest_industry_gaps.csv")
smallest_industry_gaps <- read_csv("data/smallest_industry_gaps.csv")

industry_gaps <- 
largest_industry_gaps |>
  full_join(smallest_industry_gaps, largest_industry_gaps, by = c("Occupation", "Women's Earnings as a Percentage of Men's Earnings")) |>
  arrange(desc(`Women's Earnings as a Percentage of Men's Earnings`)) |>
  clean_names()

```























