---
title: "inst_627_suarez_final"
output: html_document
date: "2023-11-7"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is the R Markdown file for Pablo Suarez's final project in INST627: Data Analytics for Information Professionals.

In this file, I am analyzing gender pay disparities in the United States.
As mentioned in my project proposal, this is an issue that has persisted for decades despite intervention from legislators and recognition among the general public. U.S. Census data suggests that it wasn't until 1990 that women reached a threshold of earning 70 cents to every dollar earned by men.
Almost 30 years later, the gap has closed to roughly 84 cents for every dollar men earn.

To better understand this issue, I would like to answer the following research questions: 
1.How does the gender pay disparity vary by race, age and education?
2.Which industries have the most and least significant wage discrepancies?
3.At what rate do men and women participate in the workforce?

As I've worked on this project, I've also considered adding the following visualization tasks.
These are things that I'd like to work on depending on whether I have enough time before the project deadline: 
1.(Potential) Highlight major policies or economic events that have contributed or shaped these trends.
2.(Potential) Create an interactive map of the United States highlighting the difference by state in annual earnings between genders.

To begin the project, I loaded in libraries that I felt would be useful toward what I wanted to accomplish.

```{r}
# Turn off scientific notation
options(scipen=999)

#load libraries
library(tidyverse)
library(janitor)
library(readr)
library(lubridate)
library(refinr)
library(dplyr)
library(ggthemes)
library(ggplot2)
library(corrr)

```

***1. How does the gender pay disparity vary by race, age and education?***

To begin working on answering this first question, I called in data from the Economic Policy Institute's State of Working America Data Library.
For the purposes of this project, I'm only focusing on male and female genders.

```{r}

gender_wage_gap <- read_csv("data/gender_wage_gap.csv")
view(gender_wage_gap)

```

The first step I took was to establish a historical baseline which included calculating the difference between the 1973 and 2022 values for median and average hourly wages by men and women.
Based on this data, since 1973, men have seen their average salary increase by \$8.86 and women by \$10.55.
The difference in median hourly salary for men over the years is less dramatic.
Male median hourly wages increased by \$0.91, while women increased by \$5.64.

Despite women seeing a greater increase in average and median hourly wages, it is a moot point when acknowledging that the difference in wages between genders has persisted over time.
For example, men in 1973 had an average hourly salary of \$26.96, while women had an average of \$17.31, for a difference of \$9.65.
In 2022, men had an average hourly salary of \$35.82, while women earned \$27.86, for a difference of \$7.96.
These findings demonstrate that while hourly wages have increased over time, little progress has been made over five decades to close the gap between men and women.

It's worth mentioning that the 2020 hourly wages were the highest in the data set, however, preferred to use the most recent earnings to preserve accuracy and because economic fluctuation is to be expected when dealing with financial data.

```{r}

gender_wage_gap |>
  summarise(
    men_avg_diff = gender_wage_gap$men_average[1] - gender_wage_gap$men_average[50], #difference in hourly average wage from 2022 to 1973 for men
    men_median_diff = gender_wage_gap$men_median[1] - gender_wage_gap$men_median[50], #difference in hourly median wage from 2022 to 1973 for men
    women_avg_diff = gender_wage_gap$women_average[1] - gender_wage_gap$women_average[50], #difference in hourly average wage from 2022 to 1973 for women
    women_median_diff = gender_wage_gap$women_median[1] - gender_wage_gap$women_median[50], #difference in hourly median wage from 2022 to 1973 for women
    mw_avg_2022 = gender_wage_gap$men_average[1] - gender_wage_gap$women_average[1], #difference in hourly average wage in 2022 between men and women
    mw_avg_1973 = gender_wage_gap$men_average[50] - gender_wage_gap$women_average[50], #difference in hourly average wage in 1973 between men and women
    mw_med_2022 = gender_wage_gap$men_median[1] - gender_wage_gap$women_median[1], #difference in hourly median wage in 2022 between men and women
    mw_med_1973 = gender_wage_gap$men_median[50] - gender_wage_gap$women_median[50] #difference in hourly median wage in 1973 between men and women
    )

```

Next, I followed a similar workflow for average hourly wages among genders by race. This data set examines white, Black and Hispanic men and women, specifically.

White Men and Women:

Between 1973 and 2022, the average hourly wages of white men have increased by \$11.17 and by \$11.93 for white women.
The hourly wage gap between white men and women in 1973 was \$10.36 and is now \$9.60, a difference of \$0.76 signifying that little has changed over time.

I calculated the difference between white women and men overall in 2022, which is \-$6.32, suggesting that white women earn that amount less than men overall. I also found the difference between white women and women overall, which equaled \$1.64, signifying that white women have higher average hourly wages compared to the average for the entire gender.

```{r}

gender_wage_gap |>
  summarise(
    white_men_avg_diff = gender_wage_gap$white_men_average[1] - gender_wage_gap$white_men_average[50],
    white_women_avg_diff = gender_wage_gap$white_women_average[1] - gender_wage_gap$white_women_average[50],
    whitemw_avg_2022 = gender_wage_gap$white_men_average[1] - gender_wage_gap$white_women_average[1],
    whitemw_avg_1973 = gender_wage_gap$white_men_average[50] - gender_wage_gap$white_women_average[50],
    wwomenvsmen2022 = gender_wage_gap$white_women_average[1] - gender_wage_gap$men_average[1],
    wwomenvswomen2022 = gender_wage_gap$white_women_average[1] - gender_wage_gap$women_average[1])

```
Black Men and Women:

Between 1973 and 2022, the average hourly wages of Black men have increased by \$6.34 and by \$8.16 for women.
The hourly wage gap between Black men and women in 1973 was \$5.26 and is now \$3.44, a difference of \$1.82, which is greater than the difference between white men and women. 

The difference between Black women and men overall in 2022 is \-$11.83, suggesting that Black women earn that amount less than men overall. I also found the difference between Black women and women overall, which equaled \-$3.87, signifying the hourly average amount less that Black women earned compared to the entire gender.

```{r}

gender_wage_gap |>
  summarise(
    black_men_avg_diff = gender_wage_gap$black_men_average[1] - gender_wage_gap$black_men_average[50],
    black_women_avg_diff = gender_wage_gap$black_women_average[1] - gender_wage_gap$black_women_average[50],
    blackmw_avg_2022 = gender_wage_gap$black_men_average[1] - gender_wage_gap$black_women_average[1],
    blackmw_avg_1973 = gender_wage_gap$black_men_average[50] - gender_wage_gap$black_women_average[50],
    bwomenvsmen2022 = gender_wage_gap$black_women_average[1] - gender_wage_gap$men_average[1],
    bwomenvswomen2022 = gender_wage_gap$black_women_average[1] - gender_wage_gap$women_average[1])

```

Hispanic Men and Women:

Between 1973 and 2022, the average hourly wages of Hispanic men have increased by \$5.96 and by \$7.21 for women.
The hourly wage gap between Hispanic men and women in 1973 was \$5.28 and is now \$4.03, a difference of \$1.25, the lowest difference among all of the races that I observed. 

The difference between Hispanic women and men overall in 2022, which is \-$13.22, the largest observed margin. The difference between Hispanic women and women overall equaled \-$5.26, signifying the hourly average amount less that Hispanic women earned compared to the entire gender.

```{r}

gender_wage_gap |>
  summarise(
    hisp_men_avg_diff = gender_wage_gap$hispanic_men_average[1] - gender_wage_gap$hispanic_men_average[50],
    hisp_women_avg_diff = gender_wage_gap$hispanic_women_average[1] - gender_wage_gap$hispanic_women_average[50],
    hispmw_avg_2022 = gender_wage_gap$hispanic_men_average[1] - gender_wage_gap$hispanic_women_average[1],
    hispmw_avg_1973 = gender_wage_gap$hispanic_men_average[50] - gender_wage_gap$hispanic_women_average[50],
    hwomenvsmen2022 = gender_wage_gap$hispanic_women_average[1] - gender_wage_gap$men_average[1],
    hwomenvswomen2022 = gender_wage_gap$hispanic_women_average[1] - gender_wage_gap$women_average[1])

```

To determine how age might factor into the gender pay gap, I cleaned up in Excel a data frame from the U.S.
Bureau of Labor Statistics showing the inflation-adjusted median of usual weekly earnings, by age groups, for full-time wage and salary workers between 1979 and 2020.
The original data can be found on table 17 of the Sept. 2021 BLS report (<https://www.bls.gov/opub/reports/womens-earnings/2020/home.html>).

This data, in its entirety, is difficult to visualize given all of the different datapoints across age groups and years. However, according to the BLS, workers earn the most when they are between the ages of 35 and 54. Therefore, I focused my visualization efforts on any age groups that fell within 35 to 54 years.

Here is line chart for the peak earning ages between men and women over the 50 years I examined: https://datawrapper.dwcdn.net/5bY1b/1/

In the code below, I calculated total earnings for the year by gender, just in case I want to use it for other calculations later on including finding the percentage of an age group's median weekly earnings compared to the total earnings by gender for a specific year.

```{r}

median_weekly_earnings_age <- read_csv("data/median_weekly_earnings_age.csv") |>
clean_names()

median_weekly_earnings_age$men_total <- rowSums(median_weekly_earnings_age[, c("men_16_to_19_years", "men_20_to_24_years", "men_25_to_34_years", "men_35_to_44_years", "men_45_to_54_years", "men_55_to_64_years", "men_65_years_and_older")])

median_weekly_earnings_age$women_total <- rowSums(median_weekly_earnings_age[, c("women_16_to_19_years", "women_20_to_24_years", "women_25_to_34_years", "women_35_to_44_years", "women_45_to_54_years", "women_55_to_64_years", "women_65_years_and_older")])

median_weekly_earnings_age <- write_csv(median_weekly_earnings_age, "data/median_weekly_earnings_age.csv")

```
The final step of this first question examines how the level of education completed might impact wage earnings.

I plan on using a BLS dataset that encompasses inflation-adjusted median usual weekly earnings, by educational attainment, for full-time wage and salary workers 25 years and older between 1979 and 2020. I'm in the process of cleaning this data, but the original data can be found in table 19 of the Sept. 2021 BLS report (<https://www.bls.gov/opub/reports/womens-earnings/2020/#table-19>).

My goal for this section is to determine how greater educational attainment among women might impact wages in comparison to men. According to the BLS, "the educational attainment of women ages 25 to 64 in the labor force has risen substantially over the past half century," and in 2019, "45 percent of women ages 25 to 64 held a bachelor’s degree and higher, compared with 11 percent in 1970." 

```{r}

men_education <- read_csv("data/men_education.csv")
women_education <- read_csv("data/women_education.csv")

```

```{r}

women_education <- women_education |>
  setNames(c("year", "w_25_plus", "w_no_hs_diploma", "w_hs_no_college", "w_some_college", "w_bach_degree_plus"))

men_education <- men_education |>
  setNames(c("year", "m_25_plus", "m_no_hs_diploma", "m_hs_no_college", "m_some_college", "m_bach_degree_plus"))

education <- merge(men_education, women_education, by = "year", all.x = TRUE)
write_csv(employment_status, "data/employment_status.csv")

```

```{r}

education <- merge(men_education, women_education, by = "year", all.x = TRUE)
write_csv(education, "data/education.csv")

```

Using the education dataset that I've created, I visualized the data into two distinct line charts: One specifically for higher education degrees (https://datawrapper.dwcdn.net/kgZUX/1/) and the other for all other levels of education (https://datawrapper.dwcdn.net/HHcz6/1/). 

```{r}

education |>
  ggplot() +
  geom_point(aes(x=m_25_plus,y=w_25_plus)) +
  geom_smooth(aes(x=m_25_plus,y=w_25_plus), method="lm")


```


***2. Which industries have the most and least significant wage discrepancies?***

To answer this question, I merged two data frames from the U.S.
Department of Labor into one featuring 50 industries and professions with the highest and lowest wage disparities by gender.
This disparity was measured with women's earnings as a percentage of men's earnings.

I then uploaded the merged data frame to Datawrapper and visualized the results.
Datawrapper Link: <https://datawrapper.dwcdn.net/HJ514/1/>

```{r}

largest_industry_gaps <- read_csv("data/largest_industry_gaps.csv")
smallest_industry_gaps <- read_csv("data/smallest_industry_gaps.csv")

industry_gaps <- 
largest_industry_gaps |>
  full_join(smallest_industry_gaps, largest_industry_gaps, by = c("Occupation", "Women's Earnings as a Percentage of Men's Earnings")) |>
  arrange(desc(`Women's Earnings as a Percentage of Men's Earnings`)) |>
  clean_names()

write_csv(industry_gaps, "data/industry_gaps.csv")
```

***3.At what rate do men and women participate in the workforce?***

To answer this question, I am calling in data from the BLS' Labor Force Statistics from the Current Population Survey which provides the employment status of the civilian non-institutional population that is 16 years and over, by sex, starting in 1982. I am in the process of cleaning this data to import it into RStudio, but the original data can be found here (https://www.bls.gov/cps/cpsaat02.htm).

My goal for this section is to identify workforce participation rates to debunk a commonly held theory that women earn less due to lower workforce participation rates. The BLS notes that the COVID-19 pandemic forced many women out of the workforce entirely due to layoffs and lack of child care. In Feb. 2021, women’s labor force participation rate was 55.8% – the same rate as April 1987. However, since then women are returning to work at a higher rate than men thanks to remote and flexible work options.

```{r}

men_employment_status <- read_csv("data/men_employment_status.csv")
women_employment_status <- read_csv("data/women_employment_status.csv")

```

```{r}

men_employment_status <- men_employment_status |>
  setNames(c("year", "m_civilian_population", "m_total", "m_percent_population", "m_total_employed", "m_pct_pop_employed","m_total_unemployed","m_pct_unemployed","m_not_in_labor_force"))

women_employment_status <- women_employment_status |>
  setNames(c("year", "w_civilian_population", "w_total", "w_percent_population", "w_total_employed", "w_pct_pop_employed","w_total_unemployed","w_pct_unemployed","w_not_in_labor_force"))

```

```{r}

employment_status <- merge(men_employment_status, women_employment_status, by = "year", all.x = TRUE)

write_csv(employment_status, "data/employment_status.csv")

```




